Typing a linear \picalc

### 2020-03-09 paper/paper.tex
Use multiplicities with polarities, dummy
Write down some of the typing rules

\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{todonotes}
\usepackage{enumitem}
\setlist{parsep=0pt,listparindent=\parindent}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[many]{tcolorbox}
\bibliographystyle{plainurl}

@@

% Commands for the pi-calculus
\newcommand{\PO}{\mathbf{0}}
\newcommand{\comp}[2]{#1 \mid #2}
\newcommand{\new}[2]{(\boldsymbol{\nu} #1 #2)\,}
\newcommand{\cout}[2]{\overline{#1}\langle#2\rangle.}
\newcommand{\cin}[2]{#1(#2).}
\newcommand{\select}[2]{#1\triangleleft#2.}
\newcommand{\branch}[2]{#1\triangleright#2}

\newcommand{\subst}[3]{#1[#2/#3]}

\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}

\newcommand{\type}{\texttt}
\newcommand{\End}{\type{End}}
\newcommand{\Send}[1]{!#1.}
\newcommand{\Recv}[1]{?#1.}
\newcommand{\Select}{\oplus}
\newcommand{\Branch}{\&}
\newcommand{\dual}{\overline}

\newcommand{\reduce}{\rightarrow}

@@

\maketitle

\begin{abstract}
  We present the syntax, operational semantics, and typing rules of a \picalc{} with linear and shared types.
  We abstract over types (using \emph{shapes} to untangle usage contexts from typing contexts) and generalize the algebras on multiplicities (using indexed sets of \emph{partial commutative monoids}).
  We use leftover typing \cite{Allais2018a} to encode our typing rules in a way that propagates linearity constraints into process continuations.
  We provide framing, weakening and strengthening proofs that we then use to prove subject congruence.
  We show that the type system is stable under substitution and prove subject reduction.

@@

@@

@@

machine verified formalisation of the pi calculus

typing with leftovers applied to the pi calculus

multiple multiplicities per variable

multiple multiplicity types

full formalisation available in Agda

@@

@@

@@

@@

@@

\begin{description}
  \item[Polarities]
  \item[Multiple variables]
  \item[Vectors]
    Most general solution, $n$ possible multiplicities
\end{description}

Generalisation over indexed sets of \emph{partial commutative monoids}

given two multiplicities, the third is uniquely determined. this is better modelled through a function rather than an inductive type.

\subsection{Variable types and multiplicities}

two-layered approach: types on one hand, capabilities on the other
removing from context vs keeping in context but marking it used

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Subject reduction}

@@

Definition
Let $\Gamma \type P \boxtimes \Delta$ and $\Gamma - Delta \equiv \Xi - \phi$.
Then $\Xi \type P \boxtimes \phi$.

By defining of $\uplus$ as a total function, $\Gamma - \Delta$ is no longer
functional: $\omega - \omega$ results in any multiplicity, including $0$. Therefore
$\Xi \type P \boxtimes \phi$ would imply that such a variable cannot appear in $P$.

@@

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{todonotes}
\usepackage{enumitem}
\setlist{parsep=0pt,listparindent=\parindent}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsbsy}
\usepackage[many]{tcolorbox}
\bibliographystyle{plainurl}

% Inference rules
\usepackage{mathpartir}

% TODO notes -- TODO: remove
\usepackage{todonotes}

@@

% Processes
\newcommand{\PO}{\mathbf{0}}
\newcommand{\comp}[2]{#1 \parallel #2}
\newcommand{\new}{\boldsymbol{\nu} \,}
\newcommand{\send}[2]{#1 \, \langle#2\rangle \,}
\newcommand{\recv}[1]{#1 \, \mathbb{()} \,}

\newcommand{\subst}[3]{#1[#2/#3]}
\newcommand{\picalc}{$\pi$-calculus}
\newcommand{\Picalc}{$\pi$-Calculus}
\newcommand{\reduce}{\longrightarrow}
\newcommand{\types}[4]{#1 \propto #2 \vdash #3 \boxtimes #4}
\newcommand{\suc}{{\scriptscriptstyle 1+}}

% Types
\newcommand{\Var}{\mathrm{VAR}}
\newcommand{\Process}{\mathrm{PROCESS}}
\newcommand{\Unused}{\mathrm{UNUSED}}
\newcommand{\PreCtx}{\mathrm{PRECTX}}
\newcommand{\Ctx}{\mathrm{CTX}}
\newcommand{\Type}{\mathrm{TYPE}\;}
\newcommand{\Idx}{\mathrm{IDX}\;}
\newcommand{\Idxs}{\mathrm{IDXS}}
\newcommand{\Carrier}{\mathrm{CARRIER}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Channel}{\mathrm{CHANNEL}}
\newcommand{\Rec}{\mathrm{REC}}

@@

\maketitle
\todo{remove todo notes}

\begin{abstract}
  We present the syntax, operational semantics, and typing rules of a \picalc{} with linear and shared types.
  We use leftover typing \cite{Allais2018a} to encode our typing rules in a way that propagates linearity constraints into process continuations.
  We generalize the algebras on multiplicities using indexed sets of \emph{partial commutative monoids}, allowing the user to choose a mix of linear, affine, gradual and shared typing.
  We provide framing, weakening and strengthening proofs that we then use to prove subject congruence.
  We show that the type system is stable under substitution and prove subject reduction.

@@

The \picalc models communication.

why resource-aware typing

@@

leftover typing

@@

\begin{description}
  \item [Machine verified formalisation of the linear pi calculus]

  \item [Typing with leftovers applied to the pi calculus]

  \item [Abstraction over multiplicities]

  \item [Full formalisation available in Agda]
\end{description}

@@

\begin{figure}[h]
\begin{mathpar}
  {\mprset{fraction={===}}
    \inferrule
    {n : \N}
    {\Var_n : Set}}

  \inferrule
  {n : \N}
  {0 : \Var_n}

  \inferrule
  {x : \Var_n}
  {\suc x : \Var_{\suc n}}
\end{mathpar}
\caption{Types of size $n$}
\end{figure}

\begin{figure}[h]
\begin{mathpar}
  {\mprset{fraction={===}}
    \inferrule
    {n : \N}
    {\Process_n : Set}}
\end{mathpar}

\begin{equation*} \label{pi-calc-syntax}
\begin{split}
  \Process_n ::=& \; \PO_n                    \\
              |& \; \new{} \Process_{\suc n}          \\
              |& \; \comp{\Process_n}{\Process_n}          \\
              |& \; \recv{\Var_n}{}\Process_{\suc n} \\
              |& \; \send{\Var_n}{\Var_n}\Process_n
\end{split}
\end{equation*}
\caption{Well-scoped grammar using de Bruijn indices}
\end{figure}

@@

\begin{figure}[h]
\begin{mathpar}
  {\mprset{fraction={===}}
    \inferrule
    {P \, Q : \Process_n}
    {P \equiv_n Q : Set}}

  \inferrule
  {P \; Q \; R : \Process_n}
  {comp-assoc : \comp{P}{\comp{Q}{R}} \equiv_n \comp{\comp{P}{Q}}{R}}

  \inferrule
  {P \; Q : \Process_n}
  {comp-sym : \comp{P}{Q} \equiv_n \comp{Q}{P}}

  \inferrule
  {P : \Process_n}
  {comp-end : \comp{P}{\PO_n} \equiv_n P}

  \inferrule
  { }
  {scope-end : \new \PO_{\suc n} \equiv_n \PO_n}

  \inferrule
  {P \, Q : \Process_{\suc n} \\ uQ : \Unused_0 \, Q}
  {scope-ext : \new (\comp{P}{Q}) \equiv_n \comp{(\new P)}{lower \; 0 \; \; Q \; uQ}}

  \inferrule
  {P : \Process_{\suc \suc n}}
  {scope-comm : \new \new P \equiv_n \new \new swap \; 0 \; P}
\end{mathpar}
\caption{Structural rewriting rules}
\end{figure}

\begin{figure}[h]
\begin{mathpar}
  \mprset{flushleft}

  {\mprset{fraction={===}}
    \inferrule
    { }
    {\Rec : Set}}

  \inferrule
  { }
  {zero : \Rec}

  \inferrule
  {r : \Rec}
  {one \; r : \Rec}

  \inferrule
  {r \; s : \Rec}
  {two \; r \; s : \Rec}

  {\mprset{fraction={===}}
    \inferrule
    {P \, Q : \Process_n \\ r : \Rec}
    {P =_r Q : Set}}

  \inferrule
  {P \; Q : \Process_n \\ P \equiv_n Q}
  {struct : P =_{zero} Q}

  \inferrule
  {P \; P' : \Process_{\suc n} \\ P =_r P'}
  {cong-scope : \new P =_{one \; r} \new P'}

  \inferrule
  {P \; P' : \Process_n \\ P =_r P'}
  {cong-comp : \comp{P}{Q} =_{one \; r} \comp{P'}{Q}}

  \inferrule
  {P \; P' : \Process_{\suc n} \\ P =_r P'}
  {cong-recv : \recv{x}P =_{one \; r} \recv{x}P'}

  \inferrule
  {P \; P' : \Process_n \\ P =_r P'}
  {cong-send : \send{x}{y}P =_{one \; r} \send{x}{y}P'}

  \inferrule
  {P : \Process_n}
  {refl : P =_{zero} P}

  \inferrule
  {P \; Q : \Process_n \\ P =_r Q}
  {sym : Q =_{one \; r} P}

  \inferrule
  {P \; Q \; R : \Process_n \\ P =_r Q \\ Q =_s R}
  {trans : P =_{two \; r \; s} R}
\end{mathpar}
\caption{Structural rewriting rules lifted to a congruent equivalence relation}
\end{figure}

@@

\begin{figure}[h]
\begin{mathpar}
  {\mprset{fraction={===}}
    \inferrule
    {n : \N}
    {\Channel_n : Set}}

  \inferrule
  { }
  {nothing : \Channel_n}

  \inferrule
  {i : \Var_n}
  {just \; i : \Channel_n}

  {\mprset{fraction={===}}
    \inferrule
    {i : \Channel_n \\ P \; Q : \Process_n}
    {P \reduce_i Q : Set}}

  \inferrule
  {i \; j : \Var_n \\ P : \Process_{\suc n} \\ Q : \Process_n \\ uP : \Unused_0 \, P}
  {comm : \comp{\recv{i}P}{\send{i}{j}{Q}} \reduce_{just \; i} \comp{lower \; 0 \; \subst{P}{j}{0} \; uP}{Q}}

  \inferrule
  {P \reduce_i P'}
  {par : \comp{P}{Q} \reduce_i \comp{P'}{Q}}

  \inferrule
  {P \reduce_i Q}
  {res : \new P \reduce_{dec\; i} \new Q}

  \inferrule
  {P = P' \\ P' \reduce_i Q}
  {struct : P \reduce_i Q}
\end{mathpar}
\caption{Operational semantics indexed by reducing channel}
\end{figure}

@@

A type system with both linear and shared resources has multiplicities $0$, $1$ and $\omega$.

@@

\subsection{Contexts}

\begin{figure}[h]
\begin{mathpar}
  {\mprset{fraction={===}}
    \inferrule
    {n : \N}
    {\PreCtx_n : Set}
  }

  \inferrule
  { }
  {[] : \PreCtx_0}

  \inferrule
  {\gamma : \PreCtx_n \\ t : \Type}
  {\gamma , t : \PreCtx_{\suc n}}
\end{mathpar}
\end{figure}

\begin{figure}[h]
\begin{mathpar}
  {\mprset{fraction={===}}
    \inferrule
    {n : \N}
    {\Idxs_n : Set}}

  \inferrule
  { }
  {[] : \Idxs_0}

  \inferrule
  {\imath : \Idxs_n \\ x : \Idx}
  {\imath , x : \Idxs_{\suc n}}

  \\

  {\mprset{fraction={===}}
    \inferrule
    {\imath : \Idxs n}
    {\Ctx_\imath : Set}}

  \inferrule
  { }
  {[] : Ctx_{[]}}

  \inferrule
  {\Gamma : \Ctx_\imath \\ c : \Carrier_x}
  {\Gamma , c : \Ctx_{\imath , x}}
\end{mathpar}
\end{figure}
two-layered approach: types on one hand, capabilities on the other
removing from context vs keeping in context but marking it used

@@

\begin{figure}[h]
\begin{mathpar}
  {\mprset{fraction={===}}
    \inferrule{
      \gamma : \PreCtx_n \\
      \inferrule*{}{
        \imath : \Idxs_n \\\\
        \Gamma : \Ctx_\imath} \\
      P : \Process_n \\
      \Delta : \Ctx_\imath}
    {\types{\gamma}{\Gamma}{P}{\Delta} : Set}}

\end{mathpar}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Subject reduction}

@@

@@

Suffices to model as a series of insertions.

@@

\item [Affine types]
